Density Matrix Susceptibility

### 2018-05-28 papers/1_dmatpol/paper-r1.tex
Paper updated with sketch of introduction

@@

The electronic energy of a closed\hyp{}shell molecule $A$ is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference, or unperturbed, state that is assumed to be known.

Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
Ideally, the electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

Development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
is extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also due to the dimensionality of the problem: the quantitative
model has to accurately predict all the $n^2$ elements of the density matrix. Even relatively small errors
in a particular matrix element could accumulate tremendously when computing traces with other matrices,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Therefore, in this Work we adopt the simplistic approach: we first develop a qualitative first\hyp{}principles model 
at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the model as a set of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = \tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
When ${\bf D}$ is expressed in orthogonal basis, the density matrix is idempotent, i.e.,
 {\bf D}^2 = {\bf D} \;.
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny has shown that
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
where the auxiliary vector ${\bf v}$ is defined as
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
\begin{equation} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
  -\Tr{ 
  -\Tr{ 
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. Therefore,
in the above result and also in later course of the Paper the dipole origin has been omitted. 
%where the dipole origin was omitted for notational clarity.
If the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that

@@

does not exist because ${\bf L}_i {\bf L}_i^T$ is singular.

@@

The exact mathematical form of the generalized density matrix susceptibilities is unknown. 
The first suscebility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge and the dipole moment, 
but not higher multipole moments, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from first principles theory.

@@

To determine the optimim set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization:
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular 
and constant.
Therefore, the exact solution for the optimal parameters is given by a Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

The electric field\hyp{}induced dipole and quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{uw}{\beta} \;. 
  \end{align}
\end{subequations}
%
Quadrupole moments were subsequently transformed to the their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:final-model.General}
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 }
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta {\BM\upmu}^{(N)} - \overline{\delta {\BM\upmu}^{(N)}}\rvert^2 }
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta \widetilde{\BM\Theta}^{(N)} - \overline{\delta \widetilde{\BM\Theta}^{(N)}}\rvert^2 }
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
wherease the latter two contain crucial information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the Work, the Hartree\hyp{}Fock level was assumed and the 6-311++G** basis set
was used. For all the calculations of the density matrix polarization susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalized by using the L{\"o}wdin symmetric orthogonalization scheme.

>>>>

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play. Although the advent in the computer technology 
brought some hopes, it is still insufficient to follow up the expensive needs.
Due to this reason, it has been of particular importance nowadays to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information with the heuristic knowledge of an atomistic organisation 
on a relatively large spatial scales, typically ranging beyond few nanometers.

A class of molecular dynamics simulation techniques was developed over the years. That was possible
due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
refinings. The advent of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level 
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics. 
The size of quantum level can be extended without compromising accuracy and maintaining good performance
by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism), (ii) understanding
the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms). EFP methods are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF) or density functional theory (DFT) levels.
It is also quite difficult to generalize the embedding potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
for all cases.

In this Work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the density matrix susceptibility (DMS) tensors that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion or dispersion,
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.
We introduce the necessary theory and derive the \emph{ab initio} density matrix susceptibility model
at the HF approximation in Section II. This model serves as a basis for generalization to the 
arbitrary one\hyp{}electron density with electronic correlation which is derived in Section II. 
The implementation details in a computer code as well as the validation procedures are detailed
in Section III. We then discuss the performance of the models in Section IV based on a few small molecules
in constant and varying electric field distributions. Finally, we conclude the presentation wit a few 
remarks for a future development in Section V.

%
%Understanding intricate mechanisms that govern
%the biomachinery of life requires thorough

@@

The electronic energy of a closed\hyp{}shell molecule $A$ is given by the time\hyp{}independent Schr{\"o}dinger equation,
%
\begin{equation}
 E^{(0)} = \tbraket{\Psi^{(0)}}{\mathscr{H}^{(0)}}{\Psi^{(0)}} \;,
\end{equation}
%
where $E$ stands for the total electronic energy, $\Psi$ the electronic wavefunction and $\mathscr{H}$
the Hamiltonian operator. In the above, the superscript $(0)$
denotes the reference or unperturbed state that is assumed to be known.
Under the influence of an external electrostatic perturbation, the total electronic energy
becomes
%
\begin{equation}
 E[{\bf F}({\bf r})] = \tbraket{\Psi}{\mathscr{H}^{(0)} + \mathscr{V}({\bf F}({\bf r}))}{\Psi}
\end{equation}
%
in which ${\bf F}({\bf r})$ is a non\hyp{}uniform electric field distribution.
The electric field\hyp{}induced energy change is an unknown functional 
of the associated change in the one\hyp{}electron density,
%
\begin{equation}\label{e:rho}
 \rho({\bf r}) = 2\sum_{\alpha\beta} D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
In Eq.~\eqref{e:rho} $\phi_\alpha({\bf r})$ is a basis funtion, which can be for instance 
an atomic orbital (AO) or a molecular orbital (MO), and $D_{\alpha\beta}$ is the corresponding
one\hyp{}particle density matrix.
If perturbation\hyp{}independent basis is chosen, the electric field\hyp{}induced change in the one\hyp{}particle density
is given by
%
\begin{equation}\label{e:rho}
 \delta \rho({\bf r}) = 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \phi_\alpha({\bf r}) \phi^*_\beta({\bf r}) \;.
\end{equation}
%
Ideally, the electric field\hyp{}induced change in the ground state's one\hyp{}particle density 
follows its energy minimum of the perturbed state. In addition, all the perturbed 
excited state one\hyp{}electron densities must be excited state solutions of the Schr{\"o}dinger equation
with the exact energy\hyp{}minimized ground state.

Development of a first\hyp{}principles theory
that is capable of accurately capturing the electric field\hyp{}dependence of the one\hyp{}particle density
is extremely difficult. This is not only because of the very complicated (and generally unknown) solution
to the electronic correlation problem, but also because
a working model has to accurately predict all the $n^2$ elements of the density matrix
where $n$ is the size of basis set. Even relatively small errors
in a particular matrix element could accumulate tremendously when computing traces with other matrices,
and in turn, obstruct the evaluated properties with non\hyp{}negligible errors.
Therefore, we adopt the following approach: we first develop a qualitative first\hyp{}principles model 
at the Hartree\hyp{}Fock (HF)
level of theory, and then generalize it to the arbitrary level using the functional form that was found 
by considering HF level,
but treating the density matrix susceptibility tensors as cetrain functions of adjustable parameters further on.

@@

It seems intuitive that $\delta {\bf D}$ should 
be directly related to the
polarization\hyp{}induced multipole moments because they are
a measure of the distortion of the electron density distribution due to the external electric
field. 
%Unfortunately, such a mathematical problem is very complicated due to its high\hyp{}order tensor nature
%with respect to the unknown $\delta {\bf D}$ and no approximate analytical solutions have been proposed
%up to date. Therefore, in this paper we adopt a somewhat different
%approach that will require \emph{a priori} knowledge from training set of known cases.
Therefore, 
we start our analysis by considering the induced electronic dipole moment 
of a closed-shell molecule, %which is the first non\hyp{}vanishing moment,
%
\begin{equation} \label{e:dmu-exact}
 \delta {\BM{\upmu}}({\bf r}_Q) = 
     2 \Tr{ 
         \left[ 
              \delta {\bf D} {\mathbb{M}({\bf r}_Q)}
         \right] } \;.
\end{equation}
%
where ${\mathbb{M}({\bf r}_Q)}$ is AO\hyp{}representation
of the dipole moment operator defined with respect to origin at ${\bf r}_Q$,
%
\begin{equation}\label{e:m}
 \left[ {\mathbb{M}({\bf r}_Q)} \right]_{\alpha\beta} = \tbraket{\alpha}{{\bf r} - {\bf r}_Q}{\beta} \;.
\end{equation}
%
(In this work we shall use the special open bold font for matrices of vectors, so that
in this case ${\mathbb{M}}$ can be considered as a $n \times n$ matrix of vectors of length 3).
Note that if the total charge is conserved, the induced moments are independent on the origin.

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny has shown that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{equation} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{equation}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This quarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

Now, if the AO basis is orthogonal, Eq.~\eqref{e:dmu-4-exact.linear-approximation} can be simplified by realising that

@@

does not exist because ${\bf L}_i {\bf L}_i^T$ is singular due to the fact that usually
$n > 3$.

@@

The exact mathematical form of the generalized density matrix susceptibilities 
from Eq.~\eqref{e:final-model.General} is unknown. 
The susceptibility from Eq.~\eqref{e:final-model.HF} is valid only under the Hartree\hyp{}Fock
approximation, with a guarantee that only the overall charge (exactly) and the dipole moment (approximately,
limited by the flavour of the distributed polarizability approximation), 
but not higher multipole moments, are correctly described.
Our strategy is to treat the susceptibilities from Eq.~\eqref{e:final-model.General}
as adjustable parameters, rather than try to derive them from first principles theory.

@@

To determine the optimim set 
$
 {\bf S} = 
\begin{pmatrix}
{\bf S}^{[1]} &
{\bf S}^{[2]} &
{\bf S}^{[3]}
\end{pmatrix}^T
$, we define the loss function $Z$ 
of the three blocks of adjustable parameters
that is subject to the least\hyp{}squares minimization,
%
\begin{equation}\label{e:Z}
 Z[{\bf S}] = \sum_N^{N_{\rm max}} \left( \delta D^{(N)} - \delta \overline{D}^{(N)} \right)^2 \;.
\end{equation}
%
%where $N_{\rm max}$ is the total number of samples. 
%The linear form of Eq.~\eqref{e:final-model.General.Parameters} ensures that the Hessian 
%of $Z$
%is non\hyp{}singular and independent on the parameters. 
The Hessian of $Z$ computed with respect to the parameters is non\hyp{}singular 
and constant.
Therefore, the exact solution for the optimal parameters is given by a Newton equation
%
\begin{equation}\label{e:Newton}
 {\bf S} = -{\bf H}^{-1} \cdot {\bf g} \;,
\end{equation}
%
where ${\bf g}$ and ${\bf H}$ are the gradient vector and Hessian matrix, respectively.
%
%\begin{equation}\label{e:Newton.S}
% {\bf S} = 
%\begin{pmatrix}
%{\bf S}^{[1]} &
%{\bf S}^{[2]} &
%{\bf S}^{[3]}
%\end{pmatrix}^T
%\end{equation}
%
The explicit forms of the gradient and Hessian are given in the Appendix~\ref{a:blocks}.
Note that the dimensions of parameter space for the block 1, 2 and 3 are all
equal to $3M$.
%unperturbed molecule. This makes them perfect candidates for general effective fragment parameters.

@@

The electric field\hyp{}induced dipole and quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{uw}{\beta} \;. 
  \end{align}
\end{subequations}
%
Quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:final-model.General}
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 }
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta {\BM\upmu}^{(N)} - \overline{\delta {\BM\upmu}^{(N)}}\rvert^2 }
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta \widetilde{\BM\Theta}^{(N)} - \overline{\delta \widetilde{\BM\Theta}^{(N)}}\rvert^2 }
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
wherease the latter two contain crucial information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the Work, the Hartree\hyp{}Fock level was assumed and the 6-311++G** basis set
was used. For all the calculations of the density matrix polarization susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalized by using the L{\"o}wdin symmetric orthogonalization scheme.
